---
title: create gwas db from ukbb
author: Murray Cadzow
---


```{r setup, include=FALSE}
library(fs)
library(vroom)
library(tidyverse)
library(dbplyr)
library(RSQLite)
library(config)
library(furrr)


plan(multisession)
```


Set up the db connection
```{r connection}
Sys.setenv(R_CONFIG_ACTIVE = "gwasdb")
conf <- config::get()
con <- dbConnect(odbc::odbc(),driver = conf$driver, database = conf$database , timeout = 100)
```

File locations for ukbb results
```{r}
if(file.exists("/Volumes/")){
  prefix <- path("/Volumes/scratch/merrimanlab/murray/gwasDB/data/")
} else {
  prefix <- path("~/data/ukbb500k_gwas/")
}

ukbb_dirs <- tribble(~ name, ~path,
  "ukbb_gout",  path(prefix,"source_data", "gout"),
  "ukbb_gout_male" , path(prefix,"source_data", "gout_male"),
  "ukbb_gout_female" , path(prefix,"source_data", "gout_female"),
  "ukbb_gout_hu" , path(prefix,"source_data", "gout_hu"),
  "ukbb_gout_hu_male" , path(prefix,"source_data", "gout_hu_male"),
  "ukbb_gout_hu_female" , path(prefix,"source_data", "gout_hu_female")
)
```



```{r}
chrs <- 1L:22L
```


## Load markers

Create marker lists using the mfi files
```
mkdir -p db_load
for chr in $(seq 1 22); do awk -v chrom=$chr 'OFS="\t" {print chrom":"$3"_"$4"_"$5,$2,chrom,$3,$4,$5}' markers/ukb_mfi_chr${chr}_v3.txt > db_load/marker_ids_chr${chr}.txt ; done
```

Load markers
```{r}
load_markers <- function(chrom) {
  markers <- vroom::vroom(file = path(prefix,"db_load",paste0("marker_ids_chr",chrom,".txt")), col_names = c("kgp_id","snp","chr","pos","ref","alt"))
  # load into b37 table
  markers %>% select(kgp_id, chr,pos, ref, alt) %>% DBI::dbWriteTable(conn = con, name = "b37", value = ., append = TRUE)
  # load into marker table
  markers %>% filter(str_detect(snp, pattern = "[0-9]+:[0-9]+_[A-Z]+_[A-Z]", negate = TRUE)) %>% select(kgp_id, marker_name = snp) %>% distinct() %>% DBI::dbWriteTable(conn = con, name = "marker", value = ., append = TRUE)
  
}
```



```{r}

map(chrs, load_markers)
#marker_files <- list.files("~/data/ukbb500k_gwas/db_load", pattern = "marker_ids_chr", full.names = TRUE)
#markers <- map_dfr(marker_files[22], vroom::vroom, col_names = c("kgp_id","snp","chr","pos","ref","alt"))
```

## Load study info

Create a tibble for the different gwases with the extra information

```{r}
study_info <- tribble(~name, ~ancestry, ~model_formula, ~gwas_date, ~n, ~n_case, ~n_control, ~imputed, ~impute_ref_panel, ~summary_only, ~citation, ~url, ~xsan_path, ~comment,
                      "ukbb_gout", "European", "gout ~ age + sex + PC1:40", "2019-08-01", 332370, 7131, 325239, TRUE, "HRC + 1KGP", FALSE, "", "", "","plink options: --logistic sex --freq case-control --geno 0.1 --missing --ci 0.95 --maf 0.0001 --hwe 0.000001 --hardy --covar gout_gwas_covar.covar --covar-name Age,pc1-pc40, plink version: v1.90b6.10 64-bit (17 Jun 2019)",
                      "ukbb_gout_male", "European", "gout ~ age + PC1:40", "2019-08-01", 159361, 6584, 152777, TRUE, "HRC + 1KGP",FALSE, "", "", "", "plink options: --ci 0.95 --covar gout_gwas_covar.covar --covar-name Age,pc1-pc40 --freq case-control --geno 0.1 --hardy --hwe 0.000001 --logistic --maf 0.0001, plink version: v1.90b6.10 64-bit (17 Jun 2019)",
                      "ukbb_gout_female", "European","gout ~ age + PC1:40", "2019-08-01", 173009, 547, 172462, TRUE, "HRC + 1KGP", FALSE, "", "", "", "plink options: --ci 0.95 --covar gout_gwas_covar.covar --covar-name Age,pc1-pc40 --freq case-control --geno 0.1 --hardy --hwe 0.000001 --logistic --maf 0.0001, plink version: v1.90b6.10 64-bit (17 Jun 2019)",
                      "ukbb_gout_hu", "European","gout_hu ~ age + sex + PC1:40","2019-08-01", 34149, 7131, 27018, TRUE, "HRC + 1KGP", FALSE, "","","" ,"plink options: --ci 0.95 --covar gwas_covar.covar --covar-name Age,pc1-pc40 --freq case-control --geno 0.1 --hardy --hwe 0.000001 --logistic sex --maf 0.0001, plink version: v1.90b6.10 64-bit (17 Jun 2019)",
                      "ukbb_gout_hu_male", "European", "gout_hu ~ age + PC1:40", "2019-08-01", 29743, 6584, 23159, TRUE, "HRC + 1KGP", FALSE, "", "", "", "plink options: --ci 0.95 --covar gwas_covar.covar --covar-name Age,pc1-pc40 --freq case-control --geno 0.1 --hardy --hwe 0.000001 --logistic --maf 0.0001, plink version: v1.90b6.10 64-bit (17 Jun 2019)",
                      "ukbb_gout_hu_female", "European", "gout_hu ~ age + PC1:40", "2019-08-01", 4406, 547, 3859, TRUE, "HRC + 1KGP", FALSE, "", "", "", "plink options: --ci 0.95 --covar gwas_covar.covar --covar-name Age,pc1-pc40 --freq case-control --geno 0.1 --hardy --hwe 0.000001 --logistic --maf 0.0001, plink version: v1.90b6.10 64-bit (17 Jun 2019)"
                      )
```

Load the study info into the database
```{r}
study_info %>% DBI::dbWriteTable(con, "study", value = ., append = TRUE)
study_info <- DBI::dbReadTable(con, "study")
```


## Load in the gwas results


```{r}
# A function that will calculate the MAF from the geno column
maf_calc <- function(geno) {
  counts <- str_split(geno, "/") %>% unlist() %>% map_dbl(as.numeric)
  af <- (counts[2] + 2 * counts[1])/(2*sum(counts))
  af
}
```


Function to load the gwas and frequency/hwe data in. SNPs get filtered if info score is less than 0.3
```{r}
pivot_hwe <- function(chunk, pos){
   wide <- chunk %>% janitor::clean_names() %>% 
    mutate(maf = furrr::future_map_dbl(geno, maf_calc)) %>% # calculate allele frequencies
    pivot_wider(values_from = c(geno,o_het,e_het,p, maf), names_from = test, id_cols = c(chr, snp, a1, a2))
   wide
   #write_tsv(wide, path("~/data/ukbb500k_gwas/chr22_wide.hwe.tsv"), append = TRUE, col_names = TRUE)
}

# load in the marker table from the DB and recreate the ref and alt columns so they can be used for the joins later on
marker_table <- DBI::dbReadTable(con, "marker") %>% rename("kgp_id_marker_table" = kgp_id) %>% mutate(alt_marker_table = str_extract(kgp_id_marker_table, "_[ACGT]+$") %>% str_remove("_"), ref_marker_table = str_extract(kgp_id_marker_table, "_[ACGT]+_") %>% str_remove_all("_"))
wrangle_gwas_data <- function(chr, studyid, ukbb_dir) {
  padded_chr <- str_pad(chr, width=2, "left",pad ="0") # tsv files had a padded chr number

  message(paste0("wrangling: chr",chr," studyid ",studyid, " dir:",ukbb_dir))
  
  # read gwas
  gwas_prefix <- str_remove(study_info[study_info[["id"]] == studyid,][["name"]], "ukbb_")
  gwas <- read_tsv(path(ukbb_dir,paste0(gwas_prefix,"_gwas_chr", padded_chr,"_add_unfiltered_p.tsv"))) %>% janitor::clean_names()
  
  # read impute marker info
  impute_info <- vroom::vroom(path(prefix,"markers",paste0("ukb_mfi_chr",chr,"_v3.txt")), col_names = c("chr_pos_alleles","snp_id", "position", "ref","alt","maf", "a1","info_score")) %>% mutate(remove_snp = ifelse(info_score < 0.3, TRUE, FALSE))
  
  # read hwe info
  message("pivoting")
  # pivot wide
  #wide_hwe <- read_tsv_chunked(path(ukbb_dir,paste0("chr",padded_chr ,".hwe.tsv")), callback = DataFrameCallback$new(pivot_hwe), chunk_size = 180000, progress = TRUE) 

  hwe_all <- vroom::vroom(path(ukbb_dir,paste0("chr",padded_chr ,".hwe.tsv")), delim = "\t") %>% janitor::clean_names() %>% filter(test == "ALL", snp %in% gwas$snp) %>% mutate(maf = furrr::future_map_dbl(geno, maf_calc)) %>% select(chr, snp, a1, a2, "geno_all" = geno, "hwe_p_all" = p, "maf_all" = maf) %>% distinct()
  hwe_aff <- vroom::vroom(path(ukbb_dir,paste0("chr",padded_chr ,".hwe.tsv")), delim = "\t") %>% janitor::clean_names() %>% filter(test == "AFF",snp %in% gwas$snp) %>% mutate(maf = furrr::future_map_dbl(geno, maf_calc)) %>% select(chr, snp, a1, a2, "geno_aff" = geno, "hwe_p_aff" = p, "maf_aff" = maf) %>% distinct()
  hwe_unaff <- vroom::vroom(path(ukbb_dir,paste0("chr",padded_chr ,".hwe.tsv")), delim = "\t") %>% janitor::clean_names() %>% filter(test == "UNAFF",snp %in% gwas$snp) %>% mutate(maf = furrr::future_map_dbl(geno, maf_calc)) %>% select(chr, snp, a1, a2, "geno_unaff" = geno, "hwe_p_unaff" = p, "maf_unaff" = maf) %>% distinct()
  
  wide_hwe2 <- left_join(hwe_all, hwe_aff) %>% left_join(hwe_unaff)
  
  # filter to only have complete data based on gwas and hwe
  res_wide <- gwas %>% filter(snp %in% wide_hwe2$snp) %>% left_join(., wide_hwe2, by = c("chr","snp", "a1"))
  
  # add on the impute info data
  gwas_hwe_info <- impute_info %>% unite("snp", snp_id, chr_pos_alleles, sep = ",", remove = FALSE) %>% select(-maf, -a1) %>% right_join(res_wide, by = "snp")
  
  message("Loading qc_removed markers")
  # fill in qc_removed table
  gwas_hwe_info %>% filter(remove_snp == TRUE ) %>% 
    mutate(study_id = studyid) %>% 
    select(kgp_id = chr_pos_alleles, study_id) %>% 
    write_csv(path(prefix,"db_load",paste0("chr",chr,"_studyid",studyid,"_qc_removed.csv")))
    #DBI::dbWriteTable(con, name = "qc_removed", value = ., append = TRUE)
  
  message("Beginning data load in db")
  # load data into gwas table
  gwas_hwe_info %>% filter(!is.na(or)) %>% 
    mutate(study_id = studyid, neg_log10_p = -log10(p)) %>% 
    left_join(marker_table, by = c("snp_id" = "marker_name", "alt" = "alt_marker_table", "ref"="ref_marker_table")) %>% 
    mutate(kgp_id = ifelse(str_detect(chr_pos_alleles, "(rs)|(Aff)"), kgp_id_marker_table, chr_pos_alleles)) %>% 
    select(kgp_id,
           study_id, 
           a1, 
           a2, 
           stat = or, 
           se, neg_log10_p,  
           impute_score = info_score, 
           maf_all, 
           maf_unaff, 
           maf_aff, 
           geno_all, 
           geno_aff, 
           geno_unaff, 
           hwe_p_all, 
           hwe_p_aff, 
           hwe_p_unaff) %>% 
    write_csv(path(prefix,"db_load",paste0("chr",chr,"_studyid",studyid,"_gwas.csv")))
    #DBI::dbWriteTable(con, name = "gwas", value = ., append = TRUE)
}


load_gwas_table <- function(chr, studyid) {
  message(paste0("INSERTING GWAS: chr",chr," studyid ",studyid))
  read_csv(path(prefix,"db_load",paste0("chr",chr,"_studyid",studyid,"_gwas.csv"))) %>% 
  DBI::dbWriteTable(con, name = "gwas", value = ., append = TRUE)
}

load_qc_table <- function(chr, studyid){
  message(paste0("INSERTING QC: chr",chr," studyid ",studyid))
  read_csv(path(prefix,"db_load",paste0("chr",chr,"_studyid",studyid,"_qc_removed.csv"))) %>% 
  DBI::dbWriteTable(con, name = "no_gwas_result", value = ., append = TRUE)
}

```


Wrangle the data into the table formats
```{r}

load_list <- left_join(study_info, ukbb_dirs, by = "name") %>% select(name, id, path) %>% left_join(crossing(name = ukbb_dirs$name, chrs = chrs), by = "name")
  
load_list %>%  pmap(., ~wrangle_gwas_data(studyid = ..2, ukbb_dir = ..3 ,chr = ..4))

#map(chrs, wrangle_gwas_data, studyid = 1, ukbb_dir = "test_db/gout")
```

load the gwas table with data
```{r, eval = FALSE}
map(chrs, load_gwas_table, studyid = 1)
```

load the qc table with data
```{r, eval = FALSE}
map(chrs, load_qc_table, studyid = 1)
```


### Dev commands

```{r, eval = FALSE}
gwas <- read_tsv("~/data/ukbb500k_gwas/gout_gwas_chr22_add_unfiltered_p.tsv") %>% janitor::clean_names()
```

load impute marker info
```{r, eval = FALSE}
impute_info <- vroom::vroom("~/data/ukbb500k_gwas/ukb_mfi_chr22_v3.txt", col_names = c("chr_pos_alleles","snp_id", "position", "ref","alt","maf", "a1","info_score")) %>% mutate(remove_snp = ifelse(info_score <= 0.3, TRUE, FALSE))
```

Load in the hwe
```{r, eval = FALSE}
hwe <- vroom::vroom("~/data/ukbb500k_gwas/chr22.hwe.tsv", n_max = 100000) %>% janitor::clean_names()
```



Use pivot wider to reduce down to one row per snp
```{r, eval = FALSE}
wide_hwe <- hwe %>% 
  mutate(maf = map_dbl(geno, maf_calc)) %>% # calculate allele frequencies
  pivot_wider(values_from = c(geno,o_het,e_het,p, maf), names_from = test)
```


```{r, eval = FALSE}
res_wide <- gwas %>% filter(snp %in% wide_hwe$snp) %>% left_join(., wide_hwe, by = c("chr","snp", "a1"))
```

```{r, eval = FALSE}
gwas_hwe_info <- impute_info %>% unite("snp", snp_id, chr_pos_alleles, sep = ",", remove = FALSE) %>% select(-maf, -a1) %>% right_join(res_wide, by = "snp")
```



```{r, eval = FALSE}
# fill in qc_removed table
gwas_hwe_info %>% filter(remove_snp == TRUE ) %>% mutate(study_id = 1) %>% select(kgp_id = chr_pos_alleles, study_id)%>% DBI::dbWriteTable(con, name = "qc_removed", value = ., append = TRUE)

# load data into gwas table
gwas_hwe_info %>% head() %>% 
  mutate(study_id = 1, neg_log10_p = -log10(p)) %>% 
  select(kgp_id = chr_pos_alleles, study_id, a1, a2, stat = or, se, neg_log10_p,  impute_score = info_score, all_maf = maf_ALL, unaff_maf = maf_UNAFF, aff_maf = maf_AFF, all_geno = geno_ALL, aff_geno = geno_AFF, unaff_geno = geno_UNAFF, all_hwe_p = p_ALL, aff_hwe_p = p_AFF, unaff_hwe_p = p_UNAFF ) %>% 
  DBI::dbWriteTable(con, name = "gwas", value = ., append = TRUE)
```



fill in the markers table
```{r, eval = FALSE}
markers %>% select(kgp_id, chr,pos, ref, alt) %>% DBI::dbWriteTable(conn = con, name = "b37", value = ., append = FALSE, overwrite = TRUE)
markers %>%  select(kgp_id, snp) %>% distinct() %>% filter(str_detect(snp, pattern = "rs", negate = TRUE)) %>% filter(str_detect(snp, "^[1-9]", negate = TRUE))
```



```{r, eval = FALSE}
DBI::dbReadTable(con, "b37")
```


